### 2015-09-21 Manuscript/SW_early_life_telomeres_v9.Rmd
First edits after DSR comments

###Abstract
1. Favourable conditions early in life can confer delayed fitness advantages. However, the mechanistic basis of these 'silver spoon effects' remains poorly understood.  Telomeres - protective caps on the ends of chromosomes - shorten in response to early life stress, and early life telomere shortening has been associated with reduced later-life survival. It is thus possible that telomeres provide a 'missing link' between early-life environmental variation and later-life fitness.

@@

3. We studied how telomere length and shortening link spatiotemporal variation in early-life conditions to survival in the Seychelles warbler (*Acrocephalus sechellensis*). The Seychelles warbler system is excellently suited to addressing this question, due to the long-term longitudinal data on survival and senescence. We combined this with cross-sectional and longitudinal telomere measurements taken from over 300 birds from multiple cohorts, and analyse how variation in the early life environment within and across cohorts affects telomere dynamics and survival.

4. We show that early-life telomere length and shortening rates vary markedly across cohorts, and that this variation is related to temporal fluctuations in food availability. Individuals born in breeding seasons with high food availability have longer telomeres and reduced telomere shortening compared to individuals born in years with low food availability. Within cohorts, we found telomere length in chicks was related to growth, and to the number of helpers present in the natal territory. However, the number of helpers was not related to telomere shortening rates, nor to telomere length in fledgling or subadult birds. Finally, we found that longer telomeres and reduced telomere shortening rates in early life were associated with increased survival later in life.

@@

While it is clear that silver spoon effects can occur, we have little understanding of how and under what conditions early-life environments will affect adult fitness. Because the phenotypic consequences of an adverse environment can depend on an individual's genetic or epigenetic makeup [@Hoffman2000; @Richards2006], the later-life consequences of a good/poor start are expected to vary among individuals, populations and species. Moreover, it is not always possible to fully quantify what constitutes a good or bad environment, and any 'hidden' environmental variation may obscure relationships relationships between the early-life conditions that are measured and adult phenotypes. Indeed, the pervasiveness of silver spoon effects varies between species [@Drummond2011], cohorts [@Reid2003] and sexes [@Wilkin2009]. In order to better understand the later-life consequences of early-life experiences, we therefore need to understand, or at least be able to measure, how the environment differentially affects individuals within a population.

A key process that has been associated with adult survival is the rate at which telomere repeats are lost. Telomeres are protective caps on the ends of chromosomes which shorten with age [@Monaghan2006], and in response to oxidative damage that can occur due to environmental stress [@VonZglinicki2002]. When telomeres reach a critical shortness cells senesce  [@Campisi2003], resulting in organismal senescence and death  [@Wong2003]. These links between telomere length and senescence have inspired a great deal of recent research into telomere ecology [reviewed in @Horn2010; @Haussmann2010; @Monaghan2014]. While there is little evidence the relationship between telomere dynamics and survival is causal [@Simons2015], there is now excellent evidence that telomeres can act as biomarkers of cost in wild populations, retaining signatures of ecological stress that are otherwise difficult to detect [@Monaghan2014; @Schultner2014; @Asghar2015].

@@

There are numerous promising avenues for future research into telomere evolutionary ecology in this system and others. Here we have considered survival, but telomere length and shortening in early life may also be linked to other components of fitness [@Monaghan2014]. Reproductive senescence occurs in the Seychelles warbler [reviewed in @Hammers2015], making this system well suited to examining how telomere length affects lifetime reproductive success. A further avenue for future research will be examining, in a quantitative genetic framework, how genetic and environmental components, and their interactions, affect telomere dynamics and senescence in natural populations [@Becker2015; @Asghar2015a]. We expect that by gaining a fuller understanding of telomere dynamics in natural populations, the fields of life-history evolution and evolutionary ecology will be greatly enhanced.

>>>>

###Abstract
1. Favourable conditions early in life can confer delayed fitness advantages. However, the mechanistic basis of these 'silver spoon effects' remains poorly understood.  Telomeres - protective caps on the ends of chromosomes - shorten in response to early life stress, and early life telomere shortening is correlated with reduced later-life survival. It is thus possible that as telomeres are able to measure the impact of early life costs, they may provide a link between early-life environmental variation and later-life fitness.

@@

3. We studied how telomere length and shortening link spatiotemporal variation in early-life conditions to survival in the Seychelles warbler (*Acrocephalus sechellensis*). The Seychelles warbler system is excellently suited to addressing this question, due to the availability of long-term longitudinal survival and senescence data unconfounded by dispersal. We combined this with cross-sectional and longitudinal telomere measurements taken from over 300 birds from multiple cohorts, and analyse how variation in the early life environment within and across cohorts affects telomere dynamics and survival.

4. We show that early-life telomere length and shortening rates vary markedly across cohorts, and that this variation is related to temporal fluctuations in food availability. Individuals born in breeding seasons with high food availability have reduced telomere shortening compared to individuals born in years with low food availability. Within cohorts, we found telomere length in chicks was related to tarsus length (which likely represents chick ), and to the number of helpers present in the natal territory. However, the number of helpers was not related to telomere shortening rates, nor to telomere length in fledged birds. Finally, we found that longer telomeres and reduced telomere shortening rates in early life were associated with increased survival later in life.

@@

While it is clear that silver spoon effects can occur, we have little understanding of how and under what conditions early-life environments will affect adult fitness. Because the phenotypic consequences of an adverse environment can depend on an individual's initial  condition, and genetic or epigenetic makeup [@Hoffman2000; @Richards2006], the later-life consequences of a good or poor start are expected to vary among individuals, populations and species. Moreover, it is not always possible to fully quantify what constitutes a good or bad environment, and any 'hidden' environmental variation may obscure relationships between the early-life conditions that are measured and adult phenotypes. Indeed, the pervasiveness of silver spoon effects varies between species [@Drummond2011], cohorts [@Reid2003] and sexes [@Wilkin2009]. In order to better understand the later-life consequences of early-life experiences, we therefore need to understand, or at least be able to measure, how the environment differentially affects individuals within a population.

A key process that has been associated with adult survival is the rate at which telomere repeats are lost. Telomeres are protective caps on the ends of chromosomes which shorten with age [@Monaghan2006], and in response to oxidative stress levels which can be elevated  due to environmental factors [@VonZglinicki2002]. When telomeres reach a critical shortness cells senesce  [@Campisi2003], resulting in organismal senescence and death  [@Wong2003]. These links between telomere length and senescence have inspired a great deal of recent research into telomere ecology [reviewed in @Horn2010; @Haussmann2010; @Monaghan2014]. While there is little evidence the relationship between telomere dynamics and survival is causal [@Simons2015], there is now excellent evidence that telomeres can act as biomarkers of cost in wild populations, retaining signatures of ecological stress that are otherwise difficult to detect [@Monaghan2014; @Schultner2014; @Asghar2015].

@@

There are numerous promising avenues for future research into telomere evolutionary ecology in this system and others. Here we have considered survival, but telomere length and shortening in early life may also be linked to other components of fitness [reviewed in @Monaghan2014]. Reproductive senescence occurs in the Seychelles warbler [reviewed in @Hammers2015], making this system well suited to examining how telomere length affects lifetime reproductive success. A further avenue for future research will be examining, in a quantitative genetic framework, how genetic and environmental components, and their interactions, affect telomere dynamics and senescence in natural populations [@Becker2015; @Asghar2015a]. We expect that by gaining a fuller understanding of telomere dynamics in natural populations, the fields of life-history evolution and evolutionary ecology will be greatly enhanced.
